Are populations like a circuit? The relationship between isolation by distance and isolation by resistance

### 2018-08-27 isolation_by_coalescence.tex
minor fix
minor latex fix
re-did figures 4,5,6

\newcommand{\comdist}{\widetilde R}
\newcommand{\comdistfn}{\widetilde \mathcal{R}}

@@

We then test the method's ability to locate barriers to migration
on the $5 \times 3$ grid with asymmetric gene flow shown in Figure \ref{fig:5x3b_grid}. 
Each nonzero movement rate is determined randomly as before.
Coalescence rates are set to 1 everywhere.
Figure \ref{fig:5x3boxplots} shows posterior distributions of the movement rates
for both coalescence and commute time inference
on this graph, with noise standard deviation equal to
1/1000 of the mean coalescence time (left)
and 1/100 of the mean coalescence time (right).
Although increased noise in the data increased uncertainty in the estimated values,
coalescence time inference correctly inferred not only the location of the barrier
but also the migration rates along all other edges within a reasonable margin of error.
Commute time inference also identified the barrier,
but was much less accurate for the other values.
We also explored making the problem harder,
by first (a) dropping rows and columns from the data corresponding to populations 2, 5, 7, and 15;
and then (b) allowing a separate coalescence rate for each location,
in both cases using a noise level of 1/100.
The resulting posterior distributions are shown in Figure \ref{fig:5x3boxplots_mult_gam},
and show substantially increased error,
but only in movement rates to the removed locations.
As before, uncertainty was increased when multiple coalescence rates were allowed,
but not as strongly as before, likely because the problem is less close to being underconstrained,
having 105 equations and 59 unknowns.

@@

\begin{figure}
\centering
     \includegraphics[scale=0.8]{figs/5x3boxplots}
    \caption{Posterior distributions for values of $g$ 
    for the 5 $\times$ 3 graph with barriers 
    for each of the analysis cases.
    See Figure \ref{fig:5x3b_grid} to see movement parameter locations.
    \plr{TODO: replace bottom two panels with same thing for commute time (eg figs/5x3b100\_com.pdf);
         put this figure in the supplement}
}
    \label{fig:5x3boxplots}
\end{figure}

@@

In order to investigate this general situation,
we tested both methods on four different $4 \times 4$ graphs (shown in Figure \ref{fig:4x4_grids}):
\emph{uniform}, where all movement rates are 1.0;
\emph{symmetric}, where movement rates are symmetric and randomly generated as before;
\emph{asymmetric}, where all movement rates are random;
and \emph{biased}, where movement rates down or to the left are equal to 2.0,
and movement rates up or to the right are 0.5.

Differences between true mean coalescence times ($C$) and mean commute times ($\comdist$) 
are shown in Figure \ref{fig:4x4coalvcomH}:
they are fairly small in the uniform and symmetric cases, moderate in the asymmetric case, 
and extreme in the biased asymmetric case.
This difference suggests that commute time inference may be strongly misled 
in the asymmetric and biased situations,
but it still could be the case that the movement rates that give the best fit of commute time
to coalescence time data could be close to the actual rates use to generate the data.

@@

\begin{figure}
\centering
     \includegraphics[scale=.8]{figs/4x4coalvcomH}
    \caption{Plots of the true values of $\comdist$ vs $C$ 
    for the uniform graph, symmetric graph, the asymmetric graph,
    and the biased asymmetric graph, 
    shown with the $y=x$ line.
    \plr{TODO: make this a 2x2 grid instead of 4x1}
    See Figure \ref{fig:4x4_grids} to see overall graph structure.}
    \label{fig:4x4coalvcomH}
\end{figure}

\begin{figure}
\centering
     \includegraphics[scale=.8]{figs/4x4box}
    \caption{Boxplots of the posterior distributions for $g$ for commute time and coalescence time
    based inference for the uniform graph, symmetric graph, 
    the asymmetric graph, and the biased symmetric graph.
    See Figure \ref{fig:4x4_grids} to see movement parameter locations.}
    \label{fig:4x4box}
\end{figure}

@@

In addition to developing a method for inferring migration rates using pairwise coalescence times,
we compare the results with that of resistance distance (i.e., commute time) based inference.
Coalescence based inference typically has lower error than commute time based inference,
along with the benefit that you can also infer coalescence rates. 
The commute time plus diversity estimate, $\comdist$, 
provides a good estimate of the coalescence times, $C$,
when gene flow rates are symmetric ($G_{ij} = G_{ji}$), 
but can be very wrong when they are asymmetric, 
including extreme discrepencies 
when there is an overall bias in the direction of the asymmetry across the landscape 
(Figure \ref{fig:4x4coalvcomH}).
When inference is performed on discrete landscapes 
with varying degrees of migrational asymmetry, 
coalescence time based inference is more accurate than commute time inference in all tested cases
(Figure \ref{fig:4x4box}).
Although performance of the commute time based method is reasonable 
when gene flow rates in the underlying model are symmetric, 
it becomes worse as the gene flow rates become more asymmetric 
to the point where there is little correlation at all with the true values 
when there is biased asymmetric migration.
On the other hand, commute time inference is much less computationally intensive, 
so much larger spatial systems can be used.
If the gene flow rates are symmetric, the higher resolution allowed by commute time inference
may outweigh the loss in accuracy.
The implication of these results for the body of literature using resistance distance methods
is that one may need to use caution when interpreting results.
For example, inferring high gene flow rates between two areas by using resistance distance
could potentially mean that those were areas were actually being fed by a common source.
Therefore, it may be worthwhile to apply a coarse resolution coalescence time based method 
alongside a finer resolution resistance distance method 
in order to check for evidence of asymmetric gene flow patterns.

@@

\begin{figure}
\centering
     \includegraphics[scale=0.8]{figs/5x3boxplots}
    \caption{Posterior distributions for values of $g$ 
    for the 5 $\times$ 3 graph with barriers 
    for each of the analysis cases.
    See Figure \ref{fig:5x3b_grid} to see movement parameter locations.
    \plr{TODO: put bottom two panels on top; on the bottom put the same thing for commute time}
}
    \label{fig:5x3boxplots_mult_gam}
\end{figure}

>>>>

\newcommand{\comdist}{\widetilde{R}}
\newcommand{\comdistfn}{\widetilde{\mathcal{R}}}

@@

We then test the method's ability to locate barriers to migration
on the $5 \times 3$ grid with asymmetric gene flow shown in Figure \ref{fig:5x3b_grid}. 
Each nonzero movement rate is determined randomly as before.
Coalescence rates are set to 1 everywhere.
Figure \ref{fig:5x3b_post_coalvcom} shows posterior distributions of the movement rates
for both coalescence time and resistance distance inference
on this graph, with noise standard deviation equal to
1/1000 of the mean coalescence time (left)
and 1/100 of the mean coalescence time (right).
Although increased noise in the data increased uncertainty in the estimated values,
coalescence time inference correctly inferred not only the location of the barrier
but also the migration rates along all other edges within a reasonable margin of error.
Commute time inference also identified the barrier,
but was much less accurate for the other values.
We also explored making the problem harder,
by first (a) dropping rows and columns from the data corresponding to populations 2, 5, 7, and 15;
and then (b) allowing a separate coalescence rate for each location,
in both cases using a noise level of 1/100.
The resulting posterior distributions are shown in Figure \ref{fig:5x3boxplots_mult_gam},
and show substantially increased error,
but only in movement rates to the removed locations.
As before, uncertainty was increased when multiple coalescence rates were allowed,
but not as strongly as before, likely because the problem is less close to being underconstrained,
having 105 equations and 59 unknowns.

@@

\begin{figure}
\centering
     \includegraphics[scale=1]{figs/5x3b_post_coalvcom}
    \caption{Posterior distributions for values of $g$ 
    for the 5 $\times$ 3 graph with barriers 
    with two different noise levels with both coalescence time and resistance distance inference.
    See Figure \ref{fig:5x3b_grid} to see movement parameter locations.
    %\plr{TODO: replace bottom two panels with same thing for commute time (eg figs/5x3b100\_com.pdf);
    %     put this figure in the supplement}
}
    \label{fig:5x3b_post_coalvcom}
\end{figure}

@@

In order to investigate this general situation,
we tested both methods on four different $4 \times 4$ graphs (shown in Figure \ref{fig:4x4_grids}):
\emph{uniform}, where all movement rates are 1.0;
\emph{symmetric}, where movement rates are symmetric and randomly generated as before;
\emph{asymmetric}, where all movement rates are random;
and \emph{biased}, where movement rates down or to the left are equal to 2.0,
and movement rates up or to the right are 0.5.
For inference, we added noise with standard deviation $1/500^\text{th}$ of the mean value.

Differences between true mean coalescence times ($C$) and mean commute times ($\comdist$) 
are shown in Figure \ref{fig:RvsC}:
they are fairly small in the uniform and symmetric cases, moderate in the asymmetric case, 
and extreme in the biased asymmetric case.
This difference suggests that commute time inference may be strongly misled 
in the asymmetric and biased situations,
but it still could be the case that the movement rates that give the best fit of commute time
to coalescence time data could be close to the actual rates use to generate the data.

@@

\begin{figure}
\centering
     \includegraphics[scale=1]{figs/RvsC}
    \caption{Plots of the true values of $\comdist$ vs $C$ 
    for the uniform graph, symmetric graph, the asymmetric graph,
    and the biased asymmetric graph, 
    shown with the $y=x$ line.
    %\plr{TODO: make this a 2x2 grid instead of 4x1}
    See Figure \ref{fig:4x4_grids} to see overall graph structure.}
    \label{fig:RvsC}
\end{figure}

\begin{figure}
\centering
     \includegraphics[scale=1]{figs/4x4boxplots_paper}
    \caption{Boxplots of the posterior distributions for $g$ for commute time and coalescence time
    based inference for the uniform graph, symmetric graph, 
    the asymmetric graph, and the biased symmetric graph.
    See Figure \ref{fig:4x4_grids} to see movement parameter locations.}
    \label{fig:4x4box}
\end{figure}

@@

In addition to developing a method for inferring migration rates using pairwise coalescence times,
we compare the results with that of resistance distance (i.e., commute time) based inference.
Coalescence based inference typically has lower error than commute time based inference,
along with the benefit that you can also infer coalescence rates. 
The commute time plus diversity estimate, $\comdist$, 
provides a good estimate of the coalescence times, $C$,
when gene flow rates are symmetric ($G_{ij} = G_{ji}$), 
but can be very wrong when they are asymmetric, 
including extreme discrepencies 
when there is an overall bias in the direction of the asymmetry across the landscape 
(Figure \ref{fig:RvsC}).
When inference is performed on discrete landscapes 
with varying degrees of migrational asymmetry, 
coalescence time based inference is more accurate than commute time inference in all tested cases
(Figure \ref{fig:4x4box}).
Although performance of the commute time based method is reasonable 
when gene flow rates in the underlying model are symmetric, 
it becomes worse as the gene flow rates become more asymmetric 
to the point where there is little correlation at all with the true values 
when there is biased asymmetric migration.
On the other hand, commute time inference is much less computationally intensive, 
so much larger spatial systems can be used.
If the gene flow rates are symmetric, the higher resolution allowed by commute time inference
may outweigh the loss in accuracy.
The implication of these results for the body of literature using resistance distance methods
is that one may need to use caution when interpreting results.
For example, inferring high gene flow rates between two areas by using resistance distance
could potentially mean that those were areas were actually being fed by a common source.
Therefore, it may be worthwhile to apply a coarse resolution coalescence time based method 
alongside a finer resolution resistance distance method 
in order to check for evidence of asymmetric gene flow patterns.

@@

\begin{figure}
\centering
     \includegraphics[scale=0.8]{figs/5x3boxplots}
    \caption{Posterior distributions for values of $g$ 
    for the 5 $\times$ 3 graph with barriers 
    for each of the analysis cases with coalescence time inference.
    See Figure \ref{fig:5x3b_grid} to see movement parameter locations.
    %\plr{TODO: put bottom two panels on top; on the bottom put the same thing for commute time}
}
    \label{fig:5x3boxplots_mult_gam}
\end{figure}
